An enhancer code compendium of early human neuronal and neural crest development

### 2025-05-22 manuscript.tex
Changes S.A.

By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
recruitment. 
Enhancer code studies have been limited to specific cell types in model organisms. 
Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
facial mesenchyme enhancers. 
We provide tools for building enhancer code compendia through our genomic deep learning package, CREsted.

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
We compare this compendium to that of the head of a four weeks post conception human embryo.
Through these comparisons we assess the enhancer code paradigm.
\par

@@

\subsection*{Human neural tube organoids recapitulate early neural and neural crest development}

We generated human neural tube organoids to model early human neural and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Table S1}). 
In both systems we could identify neural progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neural progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neural differentiation states.
This includes a cluster that expresses the neural progenitor marker \textit{SOX2}, cells that express the early neural induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

@@

Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig. 1h}), an important hallmark of neural differentiating\cite{
braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neural and neural crest development.} \\
    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. 
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

@@

\subsection*{
  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neural and neural crest 
  development
}

To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neural progenitor, neural crest and neural cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$Deep$-$NeuralTube^e$", on the organoid and embryo atlases respectively. 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined as the topics from topic modeling. 
To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Table S2}).
More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

@@

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\subsection*{
  The enhancer code underlying dorso-ventral patterning of neural progenitors
}

Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal this is topic 33, 38, 36, 54 and 48 for the organoids and
topic 34, 38, 79, 88 and 58 for the embryo. 
As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the cells of the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
(\textbf{Fig. 3d}).
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3e}).\par

To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

To assess the ability of DeepNeuralTube for prioritizing enhancer with cell type-specific activity we selected a genomic
region that was predicited to be specific for ventral neural progenitors (\textbf{Fig. 3k}). Explaining this region,
reveals multiple FOX instances and a SOX instance by both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 3j}). 
Finally, we tested the activity of this region in chicken embryos, revealing that it indeed is specifically active in
the floorplate of the neural tube (\textbf{Fig. 3l}).\par

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did not as floorplate specific. Quantifying both the number of FOX instances per region and 
the affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
specific regions (\textbf{Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neural progenitors.} \\
    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neural progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. 
    (Continues next page).
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
    O. organoid. E. Embryo.
    Also see Figure S3 and S4.
}
\end{figure}

@@

\subsection*{
  The enhancer code underlying early neural differentiation.
}

>>>>

Genomic enhancer code studies have been limited to specific cell types in model organisms. 
We created a compendium of enhancer codes for all cell types of the human neural tube and neural crest. 
We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week-post conception 
human embryo. Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared 
between the organoid and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded as heterotypic clusters of 4-5 TF binding sites within one 
nucleosome depleted region, using soft syntactic rules. We also provide counter examples to this rule, with enhancers 
encoded as homotypic clusters and by dimer motifs. Finally, we validate DeepNeuralTube using enhancer reporter assays 
and by interpreting genetic variation associated with facial features.

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
in direct comparison to an enhancer compendium we generate in parallel for the head of a four-week post conception human 
embryo. Through these comparisons we assess the enhancer code paradigm and construct an enhancer annotation and code 
resource that is a consensus between the organoid model and the human embryo.\par

@@

\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}

We generated human neural tube organoids to model early human neuronal and neural crest development. 
To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig. 1a}). 
These conditions were previously shown to generate multiple neural progenitor and
neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
To characterize the organoid-derived cell types at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and scRNA-seq) 
at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par

To evaluate the biological relevance of the cell types and states obtained from the organoid, 
we also performed 10x single cell multiomics on the head of a single four-weeks post conception human embryo, 
resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
We annotated both the organoid and embryo cells based on the expression of marker genes
(\textbf{Table S1}). 
In both systems we could identify neural progenitors, early differentiating neurons,
neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par

Next, we sub-clustered the neural progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorsal-ventral patterning, evident from the expression of dorsal-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
In both the organoid and embryo system we obtained multiple early neural differentiation states.
This includes a cluster that expresses the neural progenitor marker \textit{SOX2}, cells that express the early neural induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). 
Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). 
In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par

@@

Scoring cells of both organoid and embryo origin for the expression of cell cycle marker genes, 
we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig. 1h}), an important hallmark of differentiating neurons\cite{
braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
  \caption{
    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
    (\textbf{c}) cells colored by cell type identify. 
    (Continues next page)
  }
\end{figure}
\afterpage{\clearpage}

@@

\subsection*{
  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
  development
}

To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
neural progenitor, neural crest and neural cell type-specific enhancers, 
we trained two sequence-to-function model convolutional neural networks:
"$DeepNeuralTube^o$" and "$Deep$-$NeuralTube^e$", on the organoid and embryo atlases respectively. 
The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig. 2a}). 
The classes of the models are defined as the topics from topic modeling. 
To obtain a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par

We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Fig. S2a-b})). 
Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
and auPR of 0.17 for both models (\textbf{Fig. S2a-b})). 
Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) is predicted to be specific to pre-migratory neural crest by
both models. Given that the VISTA enhancers are annotated based on spatial localization of enhancer activity by 
\textit{in situ} hybridization, and given that it is difficult to visually distinguish between the neural tube and pre-
migratory neural crest, this result is not surprising.\par

Next, we assessed the explainability of the models. Concretely we tested if the model learned known
TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Table S2}).
More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Table S2}). 
Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ learn to use previously known motifs as features to classify regions based
on cell type-specific chromatin accessibility.\par

@@

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and TFAP2 (\textbf{Fig. 2i}). 
We tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

\subsection*{
  The enhancer code underlying dorsal-ventral patterning of neural progenitors
}

Focusing on neural progenitors, we identified topics representing cells and regions of different dorsal-ventral
identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
From ventral to dorsal these are topics 33, 38, 36, 54 and 48 for the organoids and
topics 34, 38, 79, 88 and 58 for the embryo. 
As an example, we investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
\textit{SHH} is most strongly expressed in the cells of the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed differentially  accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
(\textbf{Fig. 3d}).
Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3e}).\par

To investigate the enhancer code underlying dorsal-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorsal-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorsal-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorsal-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorsal-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorsal-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

To assess the ability of DeepNeuralTube for prioritizing enhancers with cell type-specific activity we selected a genomic
region that was predicted to be specific for ventral neural progenitors (\textbf{Fig. 3k}). Explaining this region,
reveals multiple FOX instances and a SOX instance by both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 3j}). 
Finally, we tested the activity of this region in chicken embryos, revealing that it indeed is specifically active in
the floorplate of the neural tube (\textbf{Fig. 3l}).\par

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
In the FOXA2 target enhancers of the floorplate that are also bound by FOXA2 in other systems (HepG2 and A549 cell lines), 
we identified a higher number of FOXA2 sites using FIMO\cite{
grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025
} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOXA2 instances 
in specific regions (\textbf{Fig. S4c}). This suggests that FOXA2 by itself can bind to a subset of genomic regions 
independently of the cell type (i.e., indepdent of cooperating TFS), but mainly to regions with multiple sites, and 
high-affinity sites.

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying dorsal-ventral patterning of neural progenitors.} \\
    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neural progenitors colored by cell-topic probabilities in RGB color scales of dorsal-ventral topics. 
    (Continues next page).
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) 
    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorsal-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorsal-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by SFPE1 enhancer.
    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
    \textbf{f}, Code table showing average contribution score of pattern instances for each dorsal-ventral topic. Contribution scores for $DeepNeuralTube^o$ are not outlined. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorsal-ventral topics. 
    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorsal-ventral topics. 
    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
    topic 34 (bottom).
    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
    embryo (bottom).
    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
    O. organoid. E. Embryo.
    Also see Figure S3 and S4.
}
\end{figure}

@@

\subsection*{
  The enhancer code underlying early neuronal differentiation.
}
